Sensor Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems

### 2023-04-19 latex/intro.tex
Writin

\section{Introduction}\label{section:OED intro}
Experimental design is an important part of many scientific
investigations. When considering an inverse problem, one can often
specify sensor locations (e.g.\ in geophysics and oceanography
applications), certain wavelengths (e.g.\ in MRI) or wave reflections
from the ground (e.g.\ searching for oil or using a radar). Whatever
the allowed set of observations is, one should select the optimal
observations to take, in order to increase accuracy, reduce costs, or
both.

The problem of choosing the best set of measurements is known as the
problem of \emph{optimal experimental design}. Optimal designs are
found by optimizing some \emph{design criterion}. Many design criteria
exist, and in this study we focus on the Bayesian \emph{D-optimality
criterion}. The D-optimality criterion has a simple and appealing
motivation \cite{Chaloner1995}: a D-optimal design maximizes the
expected information gain (KL divergence \cite{CoverThomas91}) between
posterior and prior. For a linear model in finite dimensions, with
Gaussian prior and iid Gaussian noise, a D-optimal design minimizes
the determinant of the posterior covariance matrix\footnote{In a
frequentist setting, a D-optimal design minimizes the determinant of
the Fisher information matrix \cite[page 16]{Ucinski05}, so the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}.

Bayesian D-optimal designs for inverse problems on function spaces
were first studied by \cite{AlexanderianGloorGhattas14,
  AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design maximizes the KL divergence between posterior
and prior measures. For a linear inverse problem with Gaussian prior
and iid Gaussian noise, a D-optimal design minimizes an expression
analogous to the determinant of the posterior covariance operator,
understood as a product of its eigenvalues.
%% Some difficulties arise in the process, but remedies can be found
%% as shown in \cite{AlexanderianGloorGhattas14}.

When one computes an optimal design using the D-optimality criterion,
the optimization process sometimes results in observations that are
very similar. For example, if an observation is thought of as
measuring some function value at $\x \in \Omega \subseteq \R^d,
d=1,2,3$ (with added error) then the optimization procedure sometimes
places sensors in very close proximity to each other (as can be seen
in Figure \ref{fig:clusterization illustration}). Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
clusterization}.

Sensor clusterization is intuitively undesirable. Selecting a small
set of sensors may lead to incomplete or biased data, as certain parts
of the system or environment may be overlooked or
underrepresented. This can result in inaccurate or unreliable
estimates of the unknown parameters of interest, potentially leading
to incorrect conclusions or decisions.

%% \subsection{Related Work}
Several remedies are suggested in the literature to phenomenon of
sensor clusterization. In a frequentist and finite-dimensional
context, \cite{fedorov1996} and \cite[chapter 2.4.3]{Ucinski05}
discuss the clusterization phenomenon and suggest an approach called
clusterization-free design. In such designs, the user enforces
observation locations to be far from each other. One way to do this is
by introducing correlated errors which account for both observation
error and model error. Another method considered is imposing distance
constraints between observations. A somewhat different approach is
suggested in \cite[page 49]{fedorov2012}, where close observations are
merged --- a procedure which obviously does not avoid
clusterization. The same problem arises in time-series analysis for
pharmacokinetic experiments. The authors of \cite{hooker2009} suggest
modeling auto-correlation time in the noise model, which is equivalent
to the correlated errors mentioned above. Any of the above mentioned
approaches might serve as a remedy and push sensors away from each
other. Yet, none offers any insight as to why clusterization
occurs. Also, as better models are employed, model error is decreased
and the clusterization phenomenon will eventually reappear. While
these approaches are practical and help us avoid sensor
clusterization, they do not provide insight as to why sensors are
clustering.

Model errors were also considered for inverse problems in
\cite{attia2020, koval2020}. The former study is focused on inferring
a Quantity of Interest (QoI). The focus of the latter is reducing
forward solves, using randomized linear algebra. Both studies present
numerical techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. Neither \cite{attia2020}, nor \cite{koval2020} mention any
effect model errors can have on sensor clusterization.

\subsection{Contribution}
The current study aims to fill the gap of understanding sensor
clusterization in infinte dimensional Bayesian linear inverse
problems. We propose and study a relaxed and analytically tractable
model for D-optimal designs. Under this model, D-optimal designs are
solutions of a constrained optimization problem, formulated using
Lagrange multipliers. The Lagrange multipliers formulation allows us
to rigorously show how model error mitigates clusterization (Section
\ref{section:non vanishing}), and why designs that exhibit
clusterization are indeed optimal when no model error is considered.

As a byproduct, the Lagrange multipliers problem for no model error
results in a characterization of D-optimal designs. The key idea is
that a D-optimal design reduces uncertainties where they are highest
first. This is the conclusion of part (5) of Theorem \ref{thm:char},
which we state below and prove in Section \ref{section:vanishing}:

>>>>

\section{Introduction}\label{section:OED intro}
Experimental design is an important part of many scientific
investigations. When considering an inverse problem, one can often
specify sensor locations (e.g.\ in electricat impedance tomography
\cite{horesh2010impedance}), certain wavelengths (e.g.\ in MRI
\cite{horesh2008mri}), or location of sensors for waves reflecting
from the ground (e.g.\ searching for oil \cite{horesh2008borehole} or
using a radar). Whatever the allowed set of measurements is, one
should select the optimal observations to take, in order to increase
accuracy, reduce costs, or both.

A set of measurements is called a \emph{design}, and an \emph{optimal
design} is found by optimizing some \emph{design
criterion}. Surprisingly, optimal designs sometimes result in
measuremets that are very similar. For example consider measurements
taken by sensors that evaluate some function at $\x \in \Omega
\subseteq \R^d, d=1,2,3$. An optimal design may place sensors in very
close proximity to each other (Figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

There is a consensus among researchers that sensor clusterization is
undesirable. Some authors acknowledge the problem of sensor
clusterization and suggest remedies \cite{fedorov1996, hooker2009,
  fedorov2012, Ucinski05, neitzel2019sparse}. Others bypass the
problem by allowing sensor locations to be chosen from a coarse finite
grid \cite{alexanderian2014sparsification,
  AlexanderianPetraStadlerEtAl16, alexanderian2018efficient,
  koval2020, attia2020}, thus eliminating the possibility of sensor
clusterization. Sadly, there is no free lunch, and choosing sensor
locations from a coarse finite grid requires solving a hard
combinatorial optimization problem. The current study is the first, to
my knowledge, that aims to gain a deep understanding of sensor
clusterization.

\subsection{D-Optimal Designs}
In this study we focus on the Bayesian \emph{D-optimality} criterion
\cite{Chaloner1995}: a D-optimal design maximizes the expected
Kullback-Leibler divergence \cite{CoverThomas91} between posterior and
prior. For a linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. For a linear
model with Gaussian noise, the Bayesian criterion is just
a regularized version of the frequentist criterion
\cite{Chaloner1995}.}.

Bayesian D-optimal designs for inverse problems on function spaces
were first studied by \cite{AlexanderianGloorGhattas14,
  AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design maximizes the KL divergence between posterior
and prior measures. For a linear inverse problem with Gaussian prior
and Gaussian noise, a D-optimal design minimizes an expression
analogous to the determinant of the posterior covariance operator
\cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
below).

Considerable effort focused on computational aspects of finding
optimal designs in infinite-dimensional Bayesian inverse problems.
Attia \cite{attia2020} focused on QoI, Koval considered model
uncertainty and focused on reducing forward and adjoint solves, using
randomized linear algebra.\cite{koval2020}.

\subsection{Remedies to Sensor Clusterization}
Several remedies to sensor clusterization were suggeted through the
years, and we will now overview some. In \emph{clusterization-free
design} locations are forced to be far from each other
\cite{Ucinski05}. This is achieved in a frequentist and
finite-dimensional setting, by either imposing distance constraints
between observations, or by introducing correlated errors which
account for both observation error and model error
\cite{Ucinski05}. Merging close measurements is also possible
\cite{fedorov2012}. Sensor clusterization is mitigated in time-series
analysis for pharmacokinetic experiments by modeling auto-correlation
time in the noise \cite{hooker2009}, and this approach is similar in
spirit to imposing error correlations as in \cite{Ucinski05}.

Many authors bypass the sensor clusterization problem by restricting
measurements to a coarse finite grid \cite{koval2020,
  AlexanderianPetraStadlerEtAl16, alexanderian2021, attia2020}, but
this practice has two flaws: First, using a coarse grid requires
solving a hard combinatorial optimization problem. Second, restricting
sensors to a such a small subset of the computatoinal domain will
necessarily give rise to sub-optimal designs. Several remedies were
offered for the first problem: Stadler suggested ell0
sparsification. Others suggested other stuff. None were offered for
the second.

%% Today, including model errors are a bona-fide????  standard in the
%% inverse problems community: \cite{attia2020}, \cite{koval2020},
%% \cite{Alexanderian2021} all empower their models.  Both studies
%% present numerical techniques for finding optimal designs when model
%% error is present. Both are restricted to linear inverse problems
%% (although in the latter the authors use their method on a nonlinear
%% problem by taking a Laplace approximation for the posterior). Both
%% find an optimal design by first solving a continuous problem for
%% sensor weights. Said solution is then sparsified to give a binary
%% design. Both studies are successful in the task of Bayesian
%% inversion. Neither \cite{attia2020}, nor \cite{koval2020} mention any
%% effect model errors can have on sensor clusterization.

\subsection{Contribution}
Any of the above mentioned approaches might serve as a remedy and push
sensors away from each other. Yet, none offers any insight as to why
clusterization occurs. Also, as better models are employed, model
error is decreased and the clusterization phenomenon will eventually
reappear. While these approaches are practical and help practitioners
avoid sensor clusterization, they do not provide insight as to why
sensors are clustering.

The current study aims to fill the gap of understanding sensor
clusterization in infinte dimensional Bayesian linear inverse
problems. We propose and study a relaxed and analytically tractable
model for D-optimal designs. Under this model, D-optimal designs are
solutions of a constrained optimization problem, formulated using
Lagrange multipliers. The Lagrange multipliers formulation allows us
to rigorously show how model error mitigates clusterization (Section
\ref{section:non vanishing}), and why designs that exhibit
clusterization are indeed optimal when no model error is considered.

As a byproduct, the Lagrange multipliers problem for no model error
results in a characterization of D-optimal designs. The key idea is
that a D-optimal design reduces uncertainties where they are highest
first. This is the conclusion of part (5) of Theorem \ref{thm:char},
which we state below and prove in Section \ref{section:vanishing}.

### 2023-04-19 latex/paper.tex
Writin

\documentclass{amsart}
\input{definitions}

\numberwithin{equation}{section}

\begin{document}

\title[Clusterization Properties of D-optimal Designs]{Clusterization
  Properties of D-optimal Designs in Infinite Dimensional Bayesian
  Linear Inverse Problems}

\author{Yair Daon}
\address{Porter School of the Environment and Earth
  Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% \curraddr{}
\email{yair.daon@gmail.com}
%% \thanks{}

%    The 2010 edition of the Mathematics Subject Classification is
%    the current definitive version.
\subjclass{Primary: 
  62F15, % Statistics - bayesian inference
  35R30, % PDE - Inverse problems
  Secondary:
  28C20, % Measure and integration - set functions and measures and integrals in
  %infinite-dimensional spaces (Wiener measure, Gaussian measure, etc.
}
%% Primary: 62F15, 35R30; Secondary: 28C20
\date{\today}

\begin{abstract}
  We investigate the problem of sensor clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. Many authors seem to believe that the fact that spatial
  correlations mitigateWe suggest an analytically tractable model for D-optimal
  designs. We argue that sensor clusterization in the case of iid
  observation noise is generic. We also prove that a model error term
  mitigates the clusterization phenomenon.

@@

%% \section{Open Problems}
%% Some open problems to

%% We two phenomena for a generic inverse problem with unit-norm
%% constraints on obsevations: First, we showed that when model error is
%% present and obsevation noise is small --- sensor clusterization will
%% not occur. In contrast, we saw that when no model error is present ---
%% sensor clusterization is quite generic and poses no obstruction to
%% D-optimality.

\section{Acknowledgements}
This study is a part of my PhD thesis. Most of it was written under
the instruction of Prof. Georg Stadler in the Courant Institute of
Mathematical Sciences in New York University. I would like to thank
Prof. Stadler for his great mentorship, attention to details and kind
behavior.

@@

\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/library.bib}

>>>>

\documentclass{siamonline220329}
%\documentclass{amsart}
\input{definitions}

%\numberwithin{equation}{section}

\title{Sensor Clusterization in D-optimal Designs for Bayesian Linear
  Inverse Problems}

\author{Yair Daon\thanks{Porter School of the Environment and Earth
    Sciences, Tel Aviv University, Tel Aviv, Israel
    (\email{yair.daon@gmail.com}).}}
%% \address{Porter School of the Environment and Earth
%%   Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{}
%% \email{yair.daon@gmail.com}
% The 2010 edition of the Mathematics Subject Classification is the
% current definitive version.
%% \subjclass{Primary: 
%%   62F15, % Statistics - bayesian inference
%%   35R30, % PDE - Inverse problems
%%   Secondary:
%%   28C20,
%%   % Measure and integration - set functions and measures and integrals
%%          % in infinite-dimensional spaces (Wiener measure, Gaussian
%%   % measure, etc.
%% }

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  We investigate the problem of sensor clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. Many authors seem to believe that the fact that spatial
  correlations mitigateWe suggest an analytically tractable model for D-optimal
  designs. We argue that sensor clusterization in the case of iid
  observation noise is generic. We also prove that a model error term
  mitigates the clusterization phenomenon.

@@

%% \section{Open Problems}

\section{Acknowledgements}
This study is a part of my PhD thesis. Most of it was written under
the instruction of Prof. Georg Stadler of the Courant Institute of
Mathematical Sciences. I would like to thank Prof. Stadler for his
great mentorship, attention to details and kind behavior.

@@

\bibliographystyle{siamplain}
\bibliography{/home/yair/projects/library.bib}

### 2023-04-19 latex/prelims.tex
Writin

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, $\fwd \prcov \fwd^*$ is assumed invertible
--- an assumption which will be used below. Let $m$ the number of
observations taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly
separated.} $\obs \in ( \hilo^* )^m$, so that:

@@

D-optimal designs in infinite-dimensional Hilbert spaces were
investigated in \cite{AlexanderianGloorGhattas14}. Their main result
is summarized, in the notation presented above, in the following
theorem\footnote{In \cite{AlexanderianGloorGhattas14}, results are
stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but the
authors also mention these results hold for more general covariance
matrices (p. 681).}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

\noindent For an intuitive understanding of Definition \ref{def:d
  optimality}, note that from \eqref{eq:postcov} and
\eqref{eq:objective}:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \\
    &= \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \det \prcov \det \postcov^{-1}.
  \end{split}
\end{align*}
Since $\prcov$ is constant, a D-optimal design minimizes the posterior
covariance determinant, analogously to the finite-dimensional case.

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, $\fwd \prcov \fwd^*$ is assumed invertible
--- an assumption which will be used below. Let $m$ the number of
observations taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly
separated \cite{cvetkovic2023choosing}.} $\obs \in ( \hilo^* )^m$, so that:

@@

D-optimal designs in infinite-dimensional Hilbert spaces were
investigated in \cite{AlexanderianGloorGhattas14}. Their main result
is summarized, in the notation presented above, in the following
theorem\footnote{In \cite{AlexanderianGloorGhattas14}\footnote{A
similar theorem appears in \cite[Theorem
  1]{alexanderian2018efficient}.}, results are stated for $\Sigma=I$
(implied by $\modcov = 0,\sigma^2=1$), but the authors also mention
these results hold for more general covariance matrices (p. 681).}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

\noindent For an intuitive understanding of Definition \ref{def:d
  optimality}, note that from \eqref{eq:postcov} and
\eqref{eq:objective}:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \\
    &= \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \det \prcov \postcov^{-1}.
  \end{split}
\end{align*}
Since $\prcov$ is constant, a D-optimal design minimizes the posterior
covariance determinant, analogously to the finite-dimensional case.
